Profit Optimisation in Heterogeneous 2-link Parallel Games

### 2024-03-21 thesis/article.tex
Thesis: Adds section with bad examples (a(eps)=0, 0 a.p.)

% \usepackage{mathtools}
% \usepackage{tikz}
\usepackage{enumerate}
\usepackage{amssymb}
\usepackage{amsthm}

% Commands for wrapping properly common expressions.
\newcommand{\indeq}[1]{\stackrel{\text{#1}}{=}}
\newcommand{\RightarrowArg}[1]{\stackrel{#1}{\Longrightarrow}}
\newcommand{\LeftrightarrowArg}[1]{\stackrel{#1}{\Leftrightarrow}}
\newcommand{\NE}{\mathrm{N.E.}}
\newcommand{\as}{\mathrm{\alpha_s}}
% \newcommand{\Exp}{\mathrm{Exp}}
% \newcommand{\Expect}{{\rm I\kern-.3em E}}

% Theorem structures.
\newtheorem{definition}{Definition}[section]
\newtheorem{lemma}{Lemma}[section]

@@

Finally to acknowledge that we have arbitrarily handled any players with $\alpha(p) = \as(t)$.
Those players, if any, even though they see both edges with equal total cost, in the optimal flow $x$ they have settled in either edge $1$ or $2$.
We investigate more in depth the relation of $\as$ to the flow in [TODO: add ref].
\\[12pt]
We can now formally define $\as$ and prove some of its properties.
\begin{definition}
	Given a 2-link parallel heterogeneous game with latency functions $l_1, l_2$, we define the time-money sensitivity split function $\as: \mathbb{R}^2_+ \rightarrow (0, +\infty)$ where for given tolls $t = (t_1, t_2)$
	\[\as(t) = \frac{l_2(x_2(t)) - l_1(x_1(t))}{t_1 - t_2}\]
\end{definition}
The function is well defined since the given tolls $t$ result in a single optimal flow $x(t) = (x_1(t), x_2(t))$ [TODO: add ref].

@@

The next lemma gives us some insight of the behavior of $\as(t)$ around its edge values.
\begin{lemma}
	Given a 2-link parallel heterogeneous game with distribution function $\alpha$ and time-money sensitivity split function $\as$, the following hold:
	\begin{enumerate}[(i)]
		\item if $\alpha(0) > 0$ then $\as(t) < \alpha(0) \Rightarrow x(t) = (0, 1)$ or $(1, 0)$
		\item if $\alpha(0) = 0$ with $0$ an accumulation point, then $x_1(t), x_2(t) > 0, \forall t \in \mathbb{R}_+^2$
		\item $\lim_{|t_1 - t_2| \rightarrow +\infty}\as(t) = 0$
		\item $\lim_{t_1 - t_2 \rightarrow 0^-} \as(t) = \alpha(x_1), \lim_{t_1 - t_2 \rightarrow 0^+} \as(t) = \alpha(x_2)$, where ($x_1, x_2)$ the optimal flow under no tolls
	\end{enumerate}
\end{lemma}

\begin{proof}
	$ $
	\begin{enumerate}[(i)]
		\item Direct consequence from the fact that all players $p$ with $\as(t) < \alpha(p)$ are in the low latency edge.
		\item Since $\alpha(0) = 0$ and $0$ an accumulation point, then $\forall \delta > 0 \exists \epsilon > 0: p < \epsilon \Rightarrow \alpha(p) < \delta$.
		Therefore $\forall t \in \mathbb{R}_+^2$ if $\delta = \as(t)$ then $\exists \epsilon > 0: p < \epsilon \Rightarrow \alpha(p) < \as(t)$, which means that the lower latency edge will always have flow some $\epsilon > 0$, and therefore $x_1(t), x_2(t) > 0$.
		\item We only need to notice that $|l_1(x_1(t)) - l_2(x_2(t))| \le \max\{l_1(1), l_2(1)\} = L$ and we find
		\[\lim_{|t_1 - t_2| \rightarrow +\infty}\as(t) = \lim_{|t_1 - t_2| \rightarrow +\infty}\frac{|l_1(x_1(t)) - l_2(x_2(t))|}{|t_1 - t_2|} \le \lim_{|t_1 - t_2| \rightarrow +\infty}\frac{L}{|t_1 - t_2|} = 0\]
		and since $\as(t) > 0$ we get $\lim_{|t_1 - t_2| \rightarrow +\infty}\as(t) = 0$.
		\item As $|t_1 - t_2| \rightarrow 0$ then also $|l_1(x_1(t)) - l_2(x_2(t))| \rightarrow 0$ as the game approaches the homogeneous case.
		Then flow on the low latency edge will define the limit of $\as(t)$.
			\begin{itemize}
				\item For $t_1 > t_2$ we get lower latency edge $1$ with flow $x_1(t) \rightarrow x_1$ and therefore
				\[\lim_{t_1 - t_2 \rightarrow 0^-} \as(t) = \alpha(x_1)\]
				\item For $t_1 < t_2$ we get lower latency edge $2$ with flow $x_2(t) \rightarrow x_2$ and therefore
				\[\lim_{t_1 - t_2 \rightarrow 0^+} \as(t) = \alpha(x_2)\]
			\end{itemize}
	\end{enumerate}
\end{proof}

\subsection{Example with no $\NE$}

>>>>

% \usepackage{tikz}
\usepackage{enumerate}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathtools}

% Commands for wrapping properly common expressions.
\newcommand{\indeq}[1]{\stackrel{\text{#1}}{=}}
\newcommand{\RightarrowArg}[1]{\stackrel{#1}{\Longrightarrow}}
\newcommand{\LeftrightarrowArg}[1]{\stackrel{#1}{\Leftrightarrow}}
\newcommand{\NE}{\mathrm{N.E.}}
\newcommand{\as}{\mathrm{\alpha_s}}
\newcommand{\R}{\mathbb{R}}
\DeclareMathOperator*{\argmax}{arg\,max}
% \newcommand{\Exp}{\mathrm{Exp}}
% \newcommand{\Expect}{{\rm I\kern-.3em E}}

% Theorem structures.
\newtheorem{definition}{Definition}[section]
\newtheorem{lemma}{Lemma}[section]
\newtheorem{example}{Example}[section]

@@

Finally to acknowledge that we have arbitrarily handled any players with $\alpha(p) = \as(t)$.
Those players, if any, even though they see both edges with equal total cost, in the optimal flow $x$ they have settled in either edge $1$ or $2$.
We investigate more in depth the relation of $\as$ to the flow in [TODO: add ref].
\\[12pt]
We can now formally define $\as$ and prove some of its properties.
\begin{definition}
	Given a 2-link parallel heterogeneous game with latency functions $l_1, l_2$, we define the time-money sensitivity split function $\as: \R^2_+ \rightarrow (0, +\infty)$ where for given tolls $t = (t_1, t_2)$
	\[\as(t) = \frac{l_2(x_2(t)) - l_1(x_1(t))}{t_1 - t_2}\]
\end{definition}
The function is well defined since the given tolls $t$ result in a single optimal flow $x(t) = (x_1(t), x_2(t))$ [TODO: add ref].

@@

The next lemma gives us some insight of the behavior of $\as(t)$ around its edge values.
\begin{lemma}
	\label{lemma:split_bounds}
	Given a 2-link parallel heterogeneous game with distribution function $\alpha$ and time-money sensitivity split function $\as$, the following hold:
	\begin{enumerate}[(i)]
		\item if $\alpha(0) > 0$ then $\as(t) < \alpha(0) \Rightarrow x(t) = (0, 1)$ or $(1, 0)$
		\item if $\alpha(0) = 0$ with $0$ an accumulation point, then $x_1(t), x_2(t) > 0, \forall t \in \R_+^2$
		\item $\lim_{|t_1 - t_2| \rightarrow +\infty}\as(t) = 0$
		\item $\lim_{t_1 - t_2 \rightarrow 0^-} \as(t) = \alpha(x_1), \lim_{t_1 - t_2 \rightarrow 0^+} \as(t) = \alpha(x_2)$, where ($x_1, x_2)$ the optimal flow under no tolls
	\end{enumerate}
\end{lemma}

\begin{proof}
	$ $
	\begin{enumerate}[(i)]
		\item Direct consequence from the fact that all players $p$ with $\as(t) < \alpha(p)$ are in the low latency edge.
		\item Since $\alpha(0) = 0$ and $0$ an accumulation point, then $\forall \delta > 0 \exists \epsilon > 0: p < \epsilon \Rightarrow \alpha(p) < \delta$.
		Therefore $\forall t \in \R_+^2$ if $\delta = \as(t)$ then $\exists \epsilon > 0: p < \epsilon \Rightarrow \alpha(p) < \as(t)$, which means that the lower latency edge will always have flow some $\epsilon > 0$, and therefore $x_1(t), x_2(t) > 0$.
		\item We only need to notice that $|l_1(x_1(t)) - l_2(x_2(t))| \le \max\{l_1(1), l_2(1)\} = L$ and we find
		\[\lim_{|t_1 - t_2| \rightarrow +\infty}\as(t) = \lim_{|t_1 - t_2| \rightarrow +\infty}\frac{|l_1(x_1(t)) - l_2(x_2(t))|}{|t_1 - t_2|} \le \lim_{|t_1 - t_2| \rightarrow +\infty}\frac{L}{|t_1 - t_2|} = 0\]
		and since $\as(t) > 0$ we get $\lim_{|t_1 - t_2| \rightarrow +\infty}\as(t) = 0$.
		\item As $|t_1 - t_2| \rightarrow 0$ then also $|l_1(x_1(t)) - l_2(x_2(t))| \rightarrow 0$ as the game approaches the homogeneous case.
		Then flow on the low latency edge will define the limit of $\as(t)$.
			\begin{itemize}
				\item For $t_1 > t_2$ we get lower latency edge $1$ with flow $x_1(t) \rightarrow x_1$ and therefore
				\[\lim_{t_1 - t_2 \rightarrow 0^-} \as(t) = \alpha(x_1)\]
				\item For $t_1 < t_2$ we get lower latency edge $2$ with flow $x_2(t) \rightarrow x_2$ and therefore
				\[\lim_{t_1 - t_2 \rightarrow 0^+} \as(t) = \alpha(x_2)\]
			\end{itemize}
	\end{enumerate}
\end{proof}

\subsection{Profit maximisation and distribution function}

In this section we will examine how different distribution functions $\alpha$ can affect the profit maximisation game between the toll owners.
[TODO: add framework for examining games]
We will restrict our analysis on distribution functions such that $\alpha(0) > 0$, the reason being that we want to avoid two instances.
The first is to ensure that $\alpha(p) > 0$ when $p > 0$, while the second is that $0$ is not an accumulation point.
[TODO: maybe mention add ref who also take the first restriction as it creates games with no optimal tolls]
The following two examples demonstrate the issues that might occur if these restrictions are not met.

\begin{example}
	Consider a $2$-link parallel network pricing game with distribution function $\alpha$ for which it does not hold that $\alpha(p) > 0$ when $p > 0$.
\end{example}

The property not holding for $\alpha$ means that $\exists x_\epsilon > 0 : \alpha(x_\epsilon) = 0 \forall p < x_\epsilon$.
That $x_\epsilon$ flow will completely neglect the tolls and play the congestion game only considering the flow latency, therefore they will always select the fastest edge regardless of the toll price.
This in turn means that any toll owner can exploit it and set their toll to an arbitrary high value $T$ which will force all players with $\alpha(p) > 0$ to leave, thus making their own edge the fastest one.
Even if the edge has a latency function with a high increasing rate, there will be at least a small flow $x_{\epsilon^\prime} \le x_\epsilon$ on that edge.
Therefore the owner can achieve any arbitrary profit $P$ by setting their toll price as $T + P/\epsilon$.
Since both player have that strategy available, we conclude that no equilibrium exists in the pricing game.

\begin{example}
	\label{example:a_0_ap}
	Consider a $2$-link parallel network pricing game with $l_1(x) = x, l_2(x) = 2x$ and distribution function $\alpha(p) = p$.
\end{example}

Notice that the condition $\alpha(p) > 0$ when $p > 0$ holds for this example.
However in this case $0$ is an accumulation point, and by Lemma $\ref{lemma:split_bounds}(ii)$ it follows $x_1(t), x_2(t) > 0, \forall t \in \R_+^2$.
Then Lemma $\ref{lemma:split_basic}$ entails that, assuming $t_1 > t_2$, $\alpha(x_1(t)) \le \as(t) \le \alpha(x_1(t) + \epsilon), \forall \epsilon > 0$ arbitrarily small, and since in this example $\alpha$ is onto, then $\as(t) = \alpha(x_1(t))$.

We will calculate $B_i(t_{-i})$ by calculating $x_i(t)$ and then finding $\argmax \Pi_i(t)$.
Note that since we always have $x_1(t), x_2(t) > 0$ we don't need to take cases where the flow is on one edge.
For tolls $t = (t_1, t_2)$ and respective optimal flow $x = (x_1(t), x_2(t))$ we take the following cases.

\paragraph{If $t_1 > t_2$}
then $x_1$ is the low-latency flow and thus $\as(t) = \alpha(x_1(t)) = x_1(t)$.
We therefore get
\begin{align}
	l_1(x_1(t)) + \as(t) t_1 &= l_2(x_2(t)) + \as(t) t_2 &\Rightarrow \notag \\
	x_1(t) + x_1(t) t_1 &= 2 (1 - x_1(t)) + x_1(t) t_2 &\Rightarrow \notag \\
	x_1(t) &= \frac{2}{3 + t_1 - t_2} \label{eq:ex2_x1_1}
\end{align}
\[
	B_1^>(t_2) = \argmax_{t_2 < t_1} x_1(t) t_1 \indeq{(\ref{eq:ex2_x1_1})} \argmax_{t_2 < t_1} \frac{2 t_1}{3 + t_1 - t_2} = \argmax_{t_2 < t_1} \frac{2}{1 + \frac{3 - t_2}{t_1}} =
		\begin{cases}
			+\infty, & t_2 < 3 \\
			\textrm{undefined}, & t_2 = 3 \\
			t_2, & t_2 > 3 \\
		\end{cases}
\]
We can already see some of the problems that appear.
If $t_2 = 3$ then $x_1(t_1, 3) = 2/t_1$ and thus $\Pi_1(t_1, 3) = 2 \; \forall t_1 > 3$ resulting in $B_1^>(3)$ being undefinable.
Then for $t_2 < 3$ if we set $c = 3 - t_2 > 0$ we have $\Pi_1(t) = \frac{2}{1 + c/t_1}$ so as $t_1 \rightarrow +\infty$ we get that $\Pi_1(t)$ is increasing and $\Pi_1(t) \rightarrow 2$, making no $t_1 > t_2$ a best candidate.
Finally, if $t_2 > 3$ then $\Pi_1(t) = \frac{2}{1 - c/t_1}$ making $\Pi_1$ decrease as $t_1$ increases, and in turn making $t_2$ the best response with profit $\Pi_1(t) = {2 t_2}/3$.

Continuing for $B_2^>(t_1)$ where $x_2(t) = 1 - x_1(t) = \frac{1 + t_1 - t_2}{3 + t_1 - t_2}$, we calculate $\argmax_{t_2 < t_1} \Pi_2(t)$ by taking first order conditions and we get
\[\argmax_{t_2 < t_1} \Pi_2(t) = t_1 + 3 - \sqrt{2 t_1 + 6}\]
This value is within bounds when $t_1 + 3 - \sqrt{2 t_1 + 6} < t_1 \Rightarrow t_1 > 3/2$.
For $t_1 \le 3/2$ we have $\Pi_2(t)$ non-decreasing in $[0, t_1]$ achieving maximum at $t_2 = t_1$.
Therefore
\[
	B_2^>(t_1) =
		\begin{cases}
			t_1, & t_1 \le 3/2 \\
			t_1 + 3 - \sqrt{2 t_1 + 6}, & t_1 > 3/2
		\end{cases}
\]
with profit $\Pi_2(t) = t_1/3$ and $t_1 + 5 - 2 \sqrt{2 t_1 + 6}$ respectively.

\paragraph{If $t_1 < t_2$}
we follow a similar analysis and find the following
\begin{equation*}
	\begin{aligned}
		B_1^<(t_2) =
			\begin{cases}
				t_2, & t_2 \le 6 \\
				t_2 + 3 - \sqrt{t_2 + 3}, & t_1 > 6
			\end{cases}
	\end{aligned}
	\qquad
	\begin{aligned}
		B_2^<(t_1) =
			\begin{cases}
				+\infty, & t_1 < 3 \\
				\textrm{undefined}, & t_1 = 3 \\
				t_1, & t_1 > 3 \\
			\end{cases}
	\end{aligned}
\end{equation*}
with profits $\Pi_1(t) = {2 t_2}/3, t_2 + 4 - 2 \sqrt{t_2 + 3}$ and $\Pi_2(t) = 1 - \epsilon, 1, t_1/3$.

\paragraph{Combining} $B_1^>(t_2), B_1^<(t_2)$ and $B_2^>(t_1), B_2^<(t_1)$ by comparing at each interval the profits between the strategies $t_1 > t_2$ and $t_1 < t_2$ (i.e. comparing profits when playing a toll either higher or lower from the other toll owner), we get the final best response functions
\begin{equation*}
	\begin{aligned}
		B_1(t_2) =
			\begin{cases}
				+\infty, & t_2 < 3 \\
				\textrm{undefined}, & t_2 = 3 \\
				t_2, & 3 < t_2 < 6 \\
				t2 + 3 - \sqrt{t_2 + 3} & t_2 \ge 6
			\end{cases}
	\end{aligned}
	\qquad
	\begin{aligned}
		B_2(t_1) =
			\begin{cases}
				+\infty, & t_1 < 2 \sqrt{2} \\
				t_1 + 3 - \sqrt{2 t_1 + 6}, & t_1 \ge 2 \sqrt{2}
			\end{cases}
	\end{aligned}
\end{equation*}
Both toll owners have one thing in common, above a specific threshold $(t_2 > 3, t_1 > 2 \sqrt{2})$ they always play either the same or a lower toll than the other toll owner.
Below that threshold, which will eventually occur from any starting point, they both get divergent best responses and thus a $\NE$ cannot be defined.

\paragraph{The bad behaviors} are obviously having divergent or undefined best response functions, which we claim occur only when we allow $0$ to be an accumulation point in $\alpha$, since this allows players with arbitrarily low money-sensitivity to always stay on the expensive edge.
Assume that in Example $\ref{example:a_0_ap}$ we use $\alpha(p) = p + 1$ as a distribution function.
Then we get best response functions
